Python as a machine learning language: experiences from the \sklearn project

### 2013-06-10 paper.tex
refactor \textit{scikit-learn} to a command

\usepackage[utf8]{inputenc}
\usepackage{amsmath}
\usepackage{hyperref}
\usepackage[authoryear]{natbib}
\usepackage{xcolor}

% working title, please change
\title{Python as a machine learning language:
       experiences from the \textit{scikit-learn} project}

@@

\begin{abstract}
The \textit{scikit-learn} project is an increasingly popular open source machine
learning library designed to be simple and efficient, accessible to non-experts,
and reusable in various contexts. In this paper, we present and discuss our
design choices for the application programming interface (API) of the project.
In particular, we describe the simple and limited interface shared by all
learning and processing units in the library and then discuss its advantages in
terms of composition and reusability. The paper also comments on implementation
details specific to the Python ecosystem and highlights obstacles faced by users
and developers of the library.
\end{abstract}

@@

\section{The \textit{scikit-learn} project}

The \textit{scikit-learn} project~\citep{pedregosa2011} is an open source
machine learning library for the Python programming language. The ambition of
the project is to provide efficient and well-established machine learning tools
within a programming environment that is accessible to non-machine learning
experts and reusable in various scientific areas. In itself the project isn't
a novel domain-specific language, but a library that provides machine
learning idioms to a general-purpose high-level language. Among other things,
it includes classical learning algorithms, model selection tools and
preprocessing
procedures. From its core, the library has been designed to tie in with the set
of numeric and scientific packages centered around the Numpy and
SciPy~\citep{vanderwalt2011, varoquaux2013scipy} libraries. As such, \textit
{scikit-learn} is intended to be used within Python scripts and requires some
Python programming knowledge. At the time, it offers no command line or
graphical user interface.
% XXX Gael: maybe we should say that it augments a general purpose
% language with machine learning concepts, like numpy augments it with
% array-computing concepts.

Started in 2009, \textit{scikit-learn} is developed by a large and
international team of volunteers, mostly researchers from
various fields (e.g., machine learning, natural language processing, or
neuroscience). The project also greatly benefits from many occasional
contributors proposing small code or documentation fixes or improvements.
Because of this large base of developers, emphasis is put on keeping
the project maintainable. In particular, code must follow specific
quality guidelines, such as style consistency and unit-test coverage. In
addition, complete documentation and examples are required for all
features. Finally, any addition to the code base goes through rigorous
code review.

@@

In this section, we discuss the general principles of the application
programming interface (API) of \textit{scikit-learn}. We first specify the
representation of data and then proceed by describing the interface shared by
all objects of the library. We then discuss the composition and reusability
advantages of that interface.

@@

In \textit{scikit-learn}, we chose a representation of data that is as close as
possible to the matrix representation: datasets are encoded as NumPy
multidimensional arrays for dense data and as SciPy sparse matrices for sparse
data. While those may seem rather unsophisticated data representations when
compared to more object-oriented constructs, such as those used by
Weka~\citep{hall2009weka}, they bring the prime advantage of allowing us to rely
on all NumPy and SciPy vector operations -- often orders of magnitude faster
that the corresponding Python loops -- while allowing us at the same time to
keep the code short and readable.  This design choice has also been motivated by
the fact that, given their pervasiveness in many other scientific Python
packages, many scientific users of Python are already familiar with NumPy dense
arrays and SciPy sparse matrices, hence helping them to familiarize with the
project.  From a practical point of view, these formats also provide a bunch of
data loading and conversion tools which make them very easy to use in many
contexts. Moreover, for tasks where the inputs are text files or semi-structured
objects, we provide \textit{vectorizer} objects that efficiently convert such
data to the NumPy or SciPy formats.

@@

All objects within \textit{scikit-learn} share a uniform common basic API
consisting of three complementary interfaces: an \textit{estimator} interface
for building and fitting objects, a \textit{predictor} interface for making
predictions and a \textit{transformer} interface for converting data.
As much as possible, our design choices have been guided so as to avoid
the proliferation of framework code. We try to adopt simple conventions and
to limit to a minimum the number of methods an object must implement.

@@

In \textit{scikit-learn}, classical learning algorithms are not the only objects
to be implemented as estimators. For example, preprocessing routines (e.g.,
vectorization of text documents, scaling of attributes, etc) or feature
extraction techniques also implement the same interface. Even those estimators
that are entirely stateless and therefore do not require the \texttt{fit} method
to perform any useful work implement the estimator interface. As we will
illustrate in the next sections, this design pattern is indeed of prime
importance for consistency, composition and parameter validation reasons.

\textbf{Predictors.} The \textit{predictor} interface defines a \texttt{predict}
method taking as input some new data \texttt{X\_test} and producing as output
predictions for \texttt{X\_test}, based on the learned parameters of the
estimator (we call the input to \texttt{predict} ``\texttt{X\_test}'' in order
to emphasize that \texttt{predict} generalizes to new data; however,
\texttt{predict} can also be called on \texttt{X\_train}). In the case of
supervised learning estimators, this method typically returns the predicted
labels or values computed by the model.  Continuing with the previous example,
predicted labels for \texttt{X\_test} can be obtained using the following
snippet:
\begin{verbatim}
y_pred = clf.predict(X_test)
\end{verbatim}
Some unsupervised learning
estimators may however also implement that interface, as illustrated below:
\begin{verbatim}
from sklearn.cluster import KMeans
km = KMeans(n_clusters=10)
km.fit(X_train)
clust_pred = km.predict(X_test)
\end{verbatim}
In the above example, a \texttt{KMeans} estimator is first fit on the training
data \texttt{X\_train}. Then the \texttt{predict} method is called on the test
data \texttt{X\_test} in order to retrieve predicted cluster labels.
The \texttt{predict} method in \textit{scikit-learn} thus offers a unified way
to compute predictions with respect to unseen data.

@@

In \textit{scikit-learn}, the \texttt{fit} method always returns the estimator
it was called on, hence making method chaining possible.  Using this technique,
we can thus rewrite the \texttt{StandardScaler} example above in a single line:
\begin{verbatim} 
X_train = StandardScaler().fit(X_train).transform(X_train)
\end{verbatim} 
In addition, \texttt{fit(X\_train).transform(X\_train)} can further
be rewritten as \texttt{fit\_transform(X\_train)}.  Combining the \texttt{fit}
and \texttt{transform} methods into a single method \texttt{fit\_transform}
allows to share computations, which often makes a huge difference in practice.
In the same spirit, \textit{scikit-learn} sometimes provides the
\texttt{fit\_predict} method for combining the \texttt{fit} and \texttt{predict}
methods. This is especially useful for retrieving cluster predictions with
respect to the training data \texttt{X\_train} in clustering algorithms.

@@

A distinguishing feature of the \textit{scikit-learn} API is its ability to
compose new estimators from several base estimators. Composition mechanisms can
be used to combine typical machine learning workflows into a single object which
is itself an estimator. Through duck typing, composed
estimators can be used wherever usual estimators can be used. In particular,
they integrate within model selection routines, allowing one to optimize at once
every step of a complex workflow. Composition of estimators can be done in two
ways: either sequentially through \texttt{Pipeline} objects or in a parallel
fashion through \texttt{FeatureUnion} objects.

@@

% A distinguishing feature of \textit{scikit-learn} estimators
% is that they can be composed into \texttt{Pipeline} objects.
% Such objects combine typical machine learning workflows
% (extracting features, centering/scaling/normalization,
% feature selection, making predictions)
% in a single object that is itself an estimator,
% and whose methods cause data to flow through the pipeline.
% More importantly, though, they can be used for
% systematic, automated model selection.
% This is best explained by example.

>>>>

\usepackage[utf8]{inputenc}
\usepackage{amsmath}
\usepackage{hyperref}
\usepackage[authoryear]{natbib}
\usepackage{xcolor}
\usepackage{xspace}

\newcommand{\sklearn}{\textit{scikit-learn}\xspace}

% working title, please change
\title{Python as a machine learning language:
       experiences from the \sklearn project}

@@

\begin{abstract}
The \sklearn project is an increasingly popular open source machine
learning library designed to be simple and efficient, accessible to non-experts,
and reusable in various contexts. In this paper, we present and discuss our
design choices for the application programming interface (API) of the project.
In particular, we describe the simple and limited interface shared by all
learning and processing units in the library and then discuss its advantages in
terms of composition and reusability. The paper also comments on implementation
details specific to the Python ecosystem and highlights obstacles faced by users
and developers of the library.
\end{abstract}

@@

\section{The \sklearn project}

The \sklearn project~\citep{pedregosa2011} is an open source
machine learning library for the Python programming language. The ambition of
the project is to provide efficient and well-established machine learning tools
within a programming environment that is accessible to non-machine learning
experts and reusable in various scientific areas. In itself the project isn't
a novel domain-specific language, but a library that provides machine
learning idioms to a general-purpose high-level language. Among other things,
it includes classical learning algorithms, model selection tools and
preprocessing
procedures. From its core, the library has been designed to tie in with the set
of numeric and scientific packages centered around the Numpy and
SciPy~\citep{vanderwalt2011, varoquaux2013scipy} libraries.
As such, \sklearn is intended to be used within Python scripts
and requires some Python programming knowledge.
At the time, it offers no command line or graphical user interface.
% XXX Gael: maybe we should say that it augments a general purpose
% language with machine learning concepts, like numpy augments it with
% array-computing concepts.

Started in 2009, \sklearn is developed by a large and
international team of volunteers, mostly researchers from
various fields (e.g., machine learning, natural language processing, or
neuroscience). The project also greatly benefits from many occasional
contributors proposing small code or documentation fixes or improvements.
Because of this large base of developers, emphasis is put on keeping
the project maintainable. In particular, code must follow specific
quality guidelines, such as style consistency and unit-test coverage. In
addition, complete documentation and examples are required for all
features. Finally, any addition to the code base goes through rigorous
code review.

@@

In this section, we discuss the general principles of the application
programming interface (API) of \sklearn. We first specify the
representation of data and then proceed by describing the interface shared by
all objects of the library. We then discuss the composition and reusability
advantages of that interface.

@@

In \sklearn, we chose a representation of data that is as close as
possible to the matrix representation: datasets are encoded as NumPy
multidimensional arrays for dense data and as SciPy sparse matrices for sparse
data. While those may seem rather unsophisticated data representations when
compared to more object-oriented constructs, such as those used by
Weka~\citep{hall2009weka}, they bring the prime advantage of allowing us to rely
on all NumPy and SciPy vector operations -- often orders of magnitude faster
that the corresponding Python loops -- while allowing us at the same time to
keep the code short and readable.  This design choice has also been motivated by
the fact that, given their pervasiveness in many other scientific Python
packages, many scientific users of Python are already familiar with NumPy dense
arrays and SciPy sparse matrices, hence helping them to familiarize with the
project.  From a practical point of view, these formats also provide a bunch of
data loading and conversion tools which make them very easy to use in many
contexts. Moreover, for tasks where the inputs are text files or semi-structured
objects, we provide \textit{vectorizer} objects that efficiently convert such
data to the NumPy or SciPy formats.

@@

All objects within \sklearn share a uniform common basic API
consisting of three complementary interfaces: an \textit{estimator} interface
for building and fitting objects, a \textit{predictor} interface for making
predictions and a \textit{transformer} interface for converting data.
As much as possible, our design choices have been guided so as to avoid
the proliferation of framework code. We try to adopt simple conventions and
to limit to a minimum the number of methods an object must implement.

@@

In \sklearn, classical learning algorithms are not the only objects
to be implemented as estimators. For example, preprocessing routines (e.g.,
vectorization of text documents, scaling of attributes, etc) or feature
extraction techniques also implement the same interface. Even those estimators
that are entirely stateless and therefore do not require the \texttt{fit} method
to perform any useful work implement the estimator interface. As we will
illustrate in the next sections, this design pattern is indeed of prime
importance for consistency, composition and parameter validation reasons.

\textbf{Predictors.} The \textit{predictor} interface defines a \texttt{predict}
method taking as input some new data \texttt{X\_test} and producing as output
predictions for \texttt{X\_test}, based on the learned parameters of the
estimator (we call the input to \texttt{predict} ``\texttt{X\_test}'' in order
to emphasize that \texttt{predict} generalizes to new data; however,
\texttt{predict} can also be called on \texttt{X\_train}). In the case of
supervised learning estimators, this method typically returns the predicted
labels or values computed by the model.  Continuing with the previous example,
predicted labels for \texttt{X\_test} can be obtained using the following
snippet:
\begin{verbatim}
y_pred = clf.predict(X_test)
\end{verbatim}
Some unsupervised learning
estimators may however also implement that interface, as illustrated below:
\begin{verbatim}
from sklearn.cluster import KMeans
km = KMeans(n_clusters=10)
km.fit(X_train)
clust_pred = km.predict(X_test)
\end{verbatim}
In the above example, a \texttt{KMeans} estimator is first fit on the training
data \texttt{X\_train}. Then the \texttt{predict} method is called on the test
data \texttt{X\_test} in order to retrieve predicted cluster labels.
The \texttt{predict} method in \sklearn thus offers a unified way
to compute predictions with respect to unseen data.

@@

In \sklearn, the \texttt{fit} method always returns the estimator
it was called on, hence making method chaining possible.  Using this technique,
we can thus rewrite the \texttt{StandardScaler} example above in a single line:
\begin{verbatim} 
X_train = StandardScaler().fit(X_train).transform(X_train)
\end{verbatim} 
In addition, \texttt{fit(X\_train).transform(X\_train)} can further
be rewritten as \texttt{fit\_transform(X\_train)}.  Combining the \texttt{fit}
and \texttt{transform} methods into a single method \texttt{fit\_transform}
allows to share computations, which often makes a huge difference in practice.
In the same spirit, \sklearn sometimes provides the
\texttt{fit\_predict} method for combining the \texttt{fit} and \texttt{predict}
methods. This is especially useful for retrieving cluster predictions with
respect to the training data \texttt{X\_train} in clustering algorithms.

@@

A distinguishing feature of the \sklearn API is its ability to
compose new estimators from several base estimators. Composition mechanisms can
be used to combine typical machine learning workflows into a single object which
is itself an estimator. Through duck typing, composed
estimators can be used wherever usual estimators can be used. In particular,
they integrate within model selection routines, allowing one to optimize at once
every step of a complex workflow. Composition of estimators can be done in two
ways: either sequentially through \texttt{Pipeline} objects or in a parallel
fashion through \texttt{FeatureUnion} objects.

@@

% A distinguishing feature of \sklearn estimators
% is that they can be composed into \texttt{Pipeline} objects.
% Such objects combine typical machine learning workflows
% (extracting features, centering/scaling/normalization,
% feature selection, making predictions)
% in a single object that is itself an estimator,
% and whose methods cause data to flow through the pipeline.
% More importantly, though, they can be used for
% systematic, automated model selection.
% This is best explained by example.
